The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-10-02 thesis/data.tex
discussion revisions

As discussed in Section~\ref{sec:exponential_profiles}, the projected
luminosity/stellar mass density profile of dwarf galaxies is generally
well-described by an exponential law (Eq.~\ref{eq:exponential_law}). A
prototypical example is the Fornax dSph --- well-fit by an exponential
over 4.5 decades in surface density. On the other hand, dSphs like
Sculptor or Ursa Minor have profiles which deviate significantly from an
exponential profile fitted to the inner regions, with a clear excess of
stars/light in the outer regions. In this chapter, we critically review
the density profile of the MW classical dwarfs. This thesis concerns the
origin of this excess, with a tidal interpretation explored in
subsequent chapters.

@@

For a satellite, the \emph{spatial likelihood} is specified either as a
one- or two-component elliptical exponential profile. A second component
is included only when the preferred amplitude of the second component is
non-zero. The \emph{proper-motion likelihood} quantifies the agreement
of a star's motion with the dwarf galaxy's systemic motion, accounting
for observational uncertainties. The \emph{CMD likelihood} measures the
consistency of a star's \emph{Gaia} \(G\) magnitude and
\(G_{\rm BP}- G_{\rm RP}\) colour with theoretical isochrones for the
galaxy. For the background likelihoods, the spatial likelihood is a
uniform distribution over the field, and the background CMD and PM
likelihoods are constructed empirically from the stars in an annulus far
from the satellite. Each likelihood is normalized as a probability
density over the respective parameter space. Appendix
\ref{sec:density_extra} discusses the details of quality cuts and the
likelihood calculation.

@@

For our fiducial sample, we adopt a minimum probability of
\(P_{\rm sat} = 0.2\). We do not filter on magnitudes explicitly, but
J+24's quality cuts typically only include stars with \(G < 21\). We use
the \(P_{\rm sat}\) values from the elliptical 2-component runs if a
galaxy shows evidence for an outer component, the 1-component run
otherwise. Most stars have \(P_{\rm  sat}\) values which are nearly 0 or
1, so the exact choice of probability threshold has little effect on the
resulting sample. Even at our relatively generous probability threshold
of 0.2, the purity remains high when validated against spectroscopic
line-of-sight (LOS) velocities (\textasciitilde90\%, J+24). We note,
however, that these purity estimates may be biased. Stars with LOS
velocities are typically brighter and, as a result, have more precise
\emph{Gaia} measurements. However, we find that our conclusions are
unchanged when limiting samples to only the brightest stars. Altogether,
the J+24 method provides a high-quality, low-contamination sample of
dwarf galaxy candidate member stars, which we will now investigate.

@@

We illustrate the likelihood-based membership selection criteria through
a progressive tightening of criteria. First, we consider a
minimally-refined sample, \textbf{all}, which only excludes stars with
poor astrometry, unreliable photometry, or inconsistent parallaxes.
Next, incorporating CMD and PM information, we use the \textbf{CMD+PM}
to test a selection method agnostic to the spatial position. This sample
includes stars where the CMD and PM combined likelihood favours
satellite membership:
\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}.
}\end{equation} Next, our \textbf{fiducial} sample,
\(P_{\rm sat} > 0.2\), includes a spatial likelihood as described in
J+24. Finally, the \textbf{RV members} (for radial velocity) sample adds
line-of-sight velocity information (see Appendix \ref{sec:rv_obs} for
details). While the latter is the best motivated membership sample, we
do not use this sample for stellar density analysis due to its
incompleteness and complex selection function.

>>>>

As discussed in Section~\ref{sec:exponential_profiles}, the projected
luminosity/stellar mass density profile of dwarf galaxies is generally
well-described by an exponential law (Eq.~\ref{eq:exponential_law}). A
prototypical example is the Fornax dSph --- well-fit by an exponential
over 4.5 decades in surface density. On the other hand, Sculptor or Ursa
Minor have profiles that deviate significantly from an exponential
profile fitted to the inner regions, with a clear excess of stars/light
in the outer regions. In this chapter, we critically review the density
profile of the MW classical dwarfs. This thesis concerns the origin of
the outer deviations from exponential profiles, with a tidal
interpretation explored in subsequent chapters.

@@

For a satellite, the \emph{spatial likelihood} is specified either as a
one- or two-component elliptical exponential profile. A second
exponential component is considered only when the preferred amplitude of
the second component is non-zero. The \emph{proper-motion likelihood}
quantifies the agreement of a star's motion with the dwarf galaxy's
systemic motion, accounting for observational uncertainties. The
\emph{CMD likelihood} measures the consistency of a star's \emph{Gaia}
\(G\) magnitude and \(G_{\rm BP}- G_{\rm RP}\) colour with theoretical
isochrones for the dwarf. For the background likelihoods, the spatial
likelihood is a uniform distribution over the field, and the background
CMD and PM likelihoods are constructed empirically from the stars in an
annulus far from the satellite. Each likelihood is normalized as a
probability density over the respective parameter space. Appendix
\ref{sec:density_extra} discusses the details of quality cuts and the
likelihood calculation.

@@

For our fiducial sample, we adopt a minimum probability of
\(P_{\rm sat} = 0.2\). We do not filter on magnitudes explicitly, but
J+24's quality cuts typically only include stars with \(G < 21\). We use
the \(P_{\rm sat}\) values from the elliptical 2-component runs if a
galaxy shows evidence for an outer component, the 1-component run
otherwise. Most stars have \(P_{\rm  sat}\) values which are nearly 0 or
1, so the exact choice of probability threshold has little effect on the
resulting sample. Even at our relatively generous probability threshold
of 0.2, the purity remains high when validated against spectroscopic
line-of-sight (LOS) velocities (\textasciitilde90\%, J+24). We note,
however, that these purity estimates may be biased. Stars with LOS
velocities are typically brighter and, as a result, have more precise
\emph{Gaia} measurements. However, we find that our conclusions are
unchanged when limiting samples to only the brightest stars. Altogether,
the J+24 method provides a high-quality, low-contamination sample of
dwarf galaxy candidate member stars, which we will now investigate in
further detail.

@@

We illustrate the likelihood-based membership selection criteria through
a progressive tightening of criteria. First, we consider a
minimally-refined sample, \textbf{all}, which only excludes stars with
poor astrometry, unreliable photometry, or inconsistent parallaxes.
Next, incorporating CMD and PM information, we use the \textbf{CMD+PM}
to test a selection method agnostic to the spatial position. This sample
includes stars where the CMD and PM combined likelihood favours
satellite membership:
\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}.
}\end{equation} Next, our \textbf{fiducial} sample,
\(P_{\rm sat} > 0.2\), includes a spatial likelihood as described in
J+24. Finally, the \textbf{RV members} (for radial velocity) sample adds
line-of-sight velocity information (see Appendix \ref{sec:rv_obs} for
details). While the latter is the highest-purity membership sample, we
do not use this sample for stellar density analysis due to its
incompleteness and complex selection function.
